---
title: "RISK FACTOR ANALYSIS OF LUNG CANCER"
author: "Enock Bereka"
date: "2025-03-23"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(vip)
library(car)
library(sjPlot)
library(emmeans)
library(gtsummary)
library(effectsize)
library(pROC)
library(report)
library(equatiomatic)
library(ggeffects)
```

## Import the dataset
```{r warning=FALSE, message=FALSE}
data = read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/survey lung cancer_045236.csv")
```

## Inspect the dataset
```{r warning=FALSE, message=FALSE}
glimpse(data)
```

## Check for missing values
```{r warning=FALSE, message=FALSE}
anyNA(data)
```

## Check for duplicates
```{r warning=FALSE, message=FALSE}
sum(duplicated(data))
```

## Data Preprocessing
## Remove duplicates
```{r warning=FALSE, message=FALSE}
data = data %>% distinct(.)
```

## Label encoding
```{r warning=FALSE, message=FALSE}
data$LUNG_CANCER = ifelse(data$LUNG_CANCER == "YES", 1, 2)
data$GENDER = as.factor(data$GENDER)
data <- data %>% mutate_if(is.numeric, as.factor)
data$AGE <- as.numeric(data$AGE)
```

## Make names
```{r warning=FALSE, message=FALSE}
colnames(data) = make.names(colnames(data))
```

## Check for outliers in Age
```{r warning=FALSE, message=FALSE}
boxplot(data$AGE)
```

## There are no outliers in Age

## Univariate analysis
## Summary statistics for age
```{r warning=FALSE, message=FALSE}
data %>% select(AGE) %>% summary(.)
```

## The mean age of respondents is 21 indicating that majority of them were young population

## Distribution analysis
## Age
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(AGE, fill = ..count..))+
  geom_histogram(bins = 15, show.legend = F)+
  scale_fill_gradient(low = "yellow", high = "red")+
  theme_bw()+
  labs(title = "Distribution of Respondents based on Age",
       x = "Age", y = "Frequency")
```

## The graph indicates a normal distribution of the respondents age.

## Gender
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(GENDER, fill = GENDER))+
  geom_bar(show.legend = F)+
  theme_bw()+
  labs(title = "Distribution of respondents based on Gender")
```

## Smoking
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(SMOKING, fill = SMOKING))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "SMOKING", labels = c("Smoker", "non smoker"))+
  labs(title = "Distribution of respondents based on Smoking Status")
```

## Yellow Fingers
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(YELLOW_FINGERS, fill = YELLOW_FINGERS))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "YELLOW_FINGERS", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Yellow Fingers")
```

## Anxiety
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(ANXIETY, fill = ANXIETY))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "ANXIETY", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Anxiety")
```

## Peer Pressure
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(PEER_PRESSURE, fill = PEER_PRESSURE))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "PEER_PRESSURE", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Peer Pressure")
```

## Chronic Disease
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(CHRONIC.DISEASE, fill = CHRONIC.DISEASE))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "CHRONIC.DISEASE", labels = c("positive", "negative"))+
  labs(title = "Distribution of respondents based on Chronic Disease")
```

## Fatigue
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(FATIGUE, fill = FATIGUE))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "FATIGUE", labels = c("Have", "dont Have"))+
  labs(title = "Distribution of respondents based on Fatigue")
```

## Allergy
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(ALLERGY, fill = ALLERGY))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "ALLERGY", labels = c("Have", "dont have"))+
  labs(title = "Distribution of respondents based on Allergy")
```

## Alcohol Consumption
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(ALCOHOL.CONSUMING, fill = ALCOHOL.CONSUMING))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "ALCOHOL.CONSUMING", labels = c("Consumers", "non consumers"))+
  labs(title = "Distribution of respondents based on Allergy")
```

## Wheezing
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(WHEEZING, fill = WHEEZING))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "WHEEZING", labels = c("wheeze", "dont wheeze"))+
  labs(title = "Distribution of respondents based on Wheezing")
```

## Coughing
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(COUGHING, fill = COUGHING))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "COUGHING", labels = c("cough", "Dont cough"))+
  labs(title = "Distribution of respondents based on Coughing")
```

## Shortness of breath
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(SHORTNESS.OF.BREATH, fill = SHORTNESS.OF.BREATH))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "SHORTNESS.OF.BREATH", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Shortness of breath")
```

## Chest Pain
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(CHEST.PAIN, fill = CHEST.PAIN))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "CHEST.PAIN", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Chest Pain")
```

## Swallowing Difficulty
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(SWALLOWING.DIFFICULTY, fill = SWALLOWING.DIFFICULTY))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "SWALLOWING.DIFFICULTY", labels = c("Yes", "No"))+
  labs(title = "Distribution of respondents based on Swallowing Difficult")
```

## Lung Cancer
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(LUNG_CANCER, fill = LUNG_CANCER))+
  geom_bar()+
  theme_bw()+
  scale_fill_discrete(name = "LUNG_CANCER", labels = c("Negative", "Positive"))+
  labs(title = "Distribution of respondents based on Lung Cancer")
```

## Bivariate analysis
## Age vs Lung cancer
```{r warning=FALSE, message=FALSE}
t.test(AGE~LUNG_CANCER, data = data)
```

## There is no significant difference in mean age across the lung cancer status indicated by a p value > 0.05

## Boxplot of the results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = LUNG_CANCER, y = AGE, fill = LUNG_CANCER))+
  geom_boxplot()+
  theme_bw()+
  scale_fill_discrete(name = "LUNG_CANCER", labels = c("Negative", "Positive"))+
  labs(title = "Age vs Lung Cancer")
```

## Gender Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$GENDER)
contingency_table
chisq.test(contingency_table)
```

## There is no association between lung cancer and gender evidenced by p value > 0.05.

## Smoking Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$SMOKING)
contingency_table
chisq.test(contingency_table)
```

## There is no significant association between smoking and lung cancer as indicated by p-value > 0.05

## Yellow Fingers Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$YELLOW_FINGERS)
contingency_table
chisq.test(contingency_table)
```

## There is a statistically significant association between yellow fingers and lung cancer as indicated by p-value < 0.05. This shows that yellow fingers is substantially connected with lung cancer.

## Anxiety Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$ANXIETY)
contingency_table
chisq.test(contingency_table)
```

## There is a statistically significant association between Anxiety and lung cancer as shown by p-value < 0.05. This shows that anxiety is closely linked with lung cancer.

## Peer Pressure Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$PEER_PRESSURE)
contingency_table
chisq.test(contingency_table)
```

## There is a significant association between peer pressure and lung cancer as indicated by p-value < 0.05.

## Chronic Disease Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$CHRONIC.DISEASE)
contingency_table
chisq.test(contingency_table)
```

## The association between chronic disease and lung cancer is statistically significant with a p-value < 0.05.

## Fatigue Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$FATIGUE)
contingency_table
chisq.test(contingency_table)
```

## There is a significant relationship between fatigue and lung cancer with a p-value < 0.05.

## Allergy Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$ALLERGY)
contingency_table
chisq.test(contingency_table)
```

## There is a significant link between allergy and lung cancer as shown by the p-value < 0.05.

## Wheezing Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$WHEEZING)
contingency_table
chisq.test(contingency_table)
```

## There is an association between wheezing and lung cancer as indicated by p-value < 0.05.

## Alcohol Consumption Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$ALCOHOL.CONSUMING)
contingency_table
chisq.test(contingency_table)
```

## There is a sigificant connection between alcohol consumption and lung cancer as indicated by a p-value < 0.05.

## Coughing Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$COUGHING)
contingency_table
chisq.test(contingency_table)
```
## There is a substantial connection between coughing and lung cancer  as shown by the p-value < 0.05.

## Shortness of breath Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$SHORTNESS.OF.BREATH)
contingency_table
chisq.test(contingency_table)
```

## There is no significant association between shortness of breath and lung cancer as shown by a p-value < 0.05.

## Swallowing Difficulty Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$SWALLOWING.DIFFICULTY)
contingency_table
chisq.test(contingency_table)
```

## There is a substantial connection between swallowing difficulty and lung cancer as indicated by a p-value < 0.05.

## Chest Pain Vs Lung Cancer
```{r warning=FALSE, message=FALSE}
contingency_table = table(data$LUNG_CANCER, data$CHEST.PAIN)
contingency_table
chisq.test(contingency_table)
```
## There is a substantial connection between chest pain and lung cancer as shown by the p-value less than 0.05.

## Multivariate analysis
## Logistic Regression
```{r warning=FALSE, message=FALSE}
model <- glm(LUNG_CANCER~.-CHEST.PAIN-SHORTNESS.OF.BREATH-WHEEZING-ANXIETY-AGE-GENDER,
             data = data, family = binomial(link = logit))
```

## The predictors with a p-value > 0.05 were removed to enhance accuracy of the model

## Check model assumption
## Goodness of fit
```{r warning=FALSE, message=FALSE}
check_predictions(model) %>% plot()
```

## The predicted data includes the observed data.

## Outliers/Influential observations
```{r warning=FALSE, message=FALSE}
check_outliers(model)
check_outliers(model) %>% plot()
```

## Multicollinearity
```{r warning=FALSE, message=FALSE}
check_collinearity(model)
```

## There is no multicollinearity in the model indicated by low correlation among the predictors.

## Uniformity of residuals
```{r warning=FALSE, message=FALSE}
check_residuals(model)
check_residuals(model) %>% plot()
```

## Visualize the models predicted probabilities
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "CHRONIC.DISEASE")
plot_model(model, type = "pred", terms = "SMOKING")
plot_model(model, type = "pred", terms = "YELLOW_FINGERS")
plot_model(model, type = "pred", terms = "PEER_PRESSURE")
plot_model(model, type = "pred", terms = "FATIGUE")
plot_model(model, type = "pred", terms = "ALLERGY")
plot_model(model, type = "pred", terms = "ALCOHOL.CONSUMING")
plot_model(model, type = "pred", terms = "COUGHING")
plot_model(model, type = "pred", terms = "SWALLOWING.DIFFICULTY")
```

## generate odds ratios for the significant predictors
```{r warning=FALSE, message=FALSE}
tbl_regression(model, exponentiate = T, pairwise_reverse = F,
               contrasts_adjust = "bonferroni",pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
```

## The odds of a smoker developing lung cancer is 75% higher compared to the non smoker. The relationship is statistically significant.
## The odds of a respondent with yellow fingers developing lung cancer is 82% higher compared to those without yellow fingers. The relationship is statistically significant.
## The odds of someone under peer pressure developing peer pressure is 83% higher compared to those without peer pressure.
## The odds of someone with chronic disease developing lung cancer is 94% higher compared to those without chronic disease.
## The odds of participants under fatigue developing lung cancer is 94% higher compared to those without fatigue.
## The odds of participants with allergy developing lung cancer is 81% higher compared to those without allergy.
## The odds of participants who consume alcohol developing lung cancer is 83% higher compared to non consumers.
## The odds of participants who cough developing lung cancer is 95% higher compared to those who do not cough.
## The chances of participants with swallowing difficulty developing lung cancer is 97% higher compared to those with no swallowing difficulty.

## Generate the effect size of the model coefficients
```{r warning=FALSE, message=FALSE}
car::Anova(model)
```

## Visualize the models feature importance
```{r warning=FALSE, message=FALSE}
vip(model)
```

## Fatigue emerged to be the top determiniant of lung cancer and smoking the least according to my study.

## Generate model equation for making predictions
```{r warning=FALSE, message=FALSE}
extract_eq(model, use_coefs = TRUE)
```

## Models performance
```{r warning=FALSE, message=FALSE}
performance(model)
```

## The predictors explains a substantial variance of 56.6% in the study variable(lung cancer) making the model a better fit for making predictions.

## Accuracy of the model
```{r warning=FALSE, message=FALSE}
roc(LUNG_CANCER~fitted.values(model), data = data,
    plot = TRUE, legacy.axes = TRUE, 
    print.auc = TRUE, ci = TRUE)
```

## The model has an auc of 96.3% making it an excellent tool in classifying participants with lung cancer and those without.

## Mediation and moderation analysis
## Does Age modify the relationship between smoking and Lung cancer
```{r warning=FALSE, message=FALSE}
model_interaction <- glm(LUNG_CANCER ~ SMOKING * AGE, data = data, family = binomial)

## Visualize interaction effect
plot_model(model_interaction, type = "int")
```

## Does gender modify the relationship between alcohol consumption and lung cancer
```{r warning=FALSE, message=FALSE}
model_interaction <- glm(LUNG_CANCER ~ ALCOHOL.CONSUMING * GENDER, data = data, family = binomial)

## Visualize interaction effect
plot_model(model_interaction, type = "int")
```

## Does chronic disease status confound the relationship between fatigue and lung cancer
```{r warning=FALSE, message=FALSE}
model_interaction <- glm(LUNG_CANCER ~ FATIGUE * CHRONIC.DISEASE, data = data, family = binomial)

## Visualize interaction effect
plot_model(model_interaction, type = "int")
```

## Does anxiety mediate the relationship between peer pressure and lung cancer
```{r warning=FALSE, message=FALSE}
model_interaction <- glm(LUNG_CANCER ~ PEER_PRESSURE * ANXIETY, data = data, family = binomial)

## Visualize interaction effect
plot_model(model_interaction, type = "int")
```

## Is the effect of chronic disease on lung cancer mediated by fatigue
```{r warning=FALSE, message=FALSE}
model_interaction <- glm(LUNG_CANCER ~ CHRONIC.DISEASE * FATIGUE, data = data, family = binomial)

## Visualize interaction effect
plot_model(model_interaction, type = "int")
```

